The Rockerverse: Images, Packages, and Applications for Containerization with R

### 2020-01-20 manuscript.Rmd
one sentence per line

  We identified common themes across this diversity: reproducible environments, scalability and efficiency, and portability across clouds.

@@

The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by descriptions of projects developing images with R installations.
Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
We conclude with a reflection on the landscape of packages and applications and point out future directions of development.

@@

The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
This is a rather attractive proposition and Docker's very easy to use user interface has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep[e.g.,][]{Bernstein2014}, data science \citep[e.g.,][]{boettiger_introduction_2015}, and edge computing \citep[e.g.,][]{alam_orchestration_2018}.
It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
Images are normally built from plain text documents called `Dockerfile`.
A `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.

For statistical computing and analysis centered around R, the Rocker&nbsp;Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e., smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

## Images for alternative R engines
\label{alternatives}

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches, or provide features different to what the R implementation maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/) provides.
While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to others.
All these alternative implementations of R are published under GPL licenses.
Since all implementations of R claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.

**Microsoft R Open** (MRO) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and some bundled packages such as \CRANpkg{checkpoint} and specific R packages maintained by Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
Originally the optional integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread} was a distinguishing feature before it was made available publicly, and can be used in base R, too \citep{eddelbuettel_thinking_2018}.

MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g., the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability and integration into enterprise platforms.
R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf. [http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
Packages are loaded on demand, i.e., at the first call to `library()`.
Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.

@@

**FastR** ([https://github.com/oracle/fastr](https://github.com/oracle/fastr)) is "A high-performance implementation of the R programming language, built on GraalVM" \citep{oracle_labs_oraclefastr_2020}.
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g., [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).

@@

_Bioconductor_ ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R.
Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
_Bioconductor_'s 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio` images.
_Bioconductor_ installs packages based on the R version, and therefore uses `rocker/rstudio` and `rocker/r-ver` version tagging.
_Bioconductor_ selects the desired version of R from Rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
The images are summarized on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub ([https://github.com/Bioconductor/bioconductor_docker](https://github.com/Bioconductor/bioconductor_docker)), and [distributed through DockerHub](https://hub.docker.com/u/bioconductor).
Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific image tag.
A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
Images may be valuable in teaching, where participants pull pre-built images to avoid complicated configuration of their own computing environemnts.

_Bioconductor_ has several images in addition to 'base', specific to various areas of research.
The 'core' image installs the most commonly used _Bioconductor_ packages.
_Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
To simplify building and maintaining _Bioconductor_ images a Ruby templating engine is used for the `Dockerfile`s.
A recent audit of the _Bioconductor_ `Dockerfile`s, following best practices from the Docker website, led to a reduction in the size and number of layers.
The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
A subsequent layer that might appear to remove or overwrite the path actually masks, rather than alters, the original.
It is therefore important to clean up (e.g., cache removal) within each layer, and to avoid re-installing existing dependencies.

@@

**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
All these images can be run on Docker for Desktop on Windows 10.
New R versions are only added on demand.
The explorative rocker-win project demonstrates the variety of Windows on Docker, which is potentially confusing, e.g., because of the different base images.
The proof of concept provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R packages with system dependencies such as \CRANpkg{sf}.
The latter is quite comfortable and fast actually, compared to the default installation from source on Linux, because CRAN ships pre-compiled binaries for Windows.
However, compared to the various Linux images, the footprint of Windows images is quite large.

For Windows, just as originally for Linux, the cloud use cases drive the development and Docker and Microsoft collaborate closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
Naturally, a Windows license is required for the host machine, and the licensing impacts the potential to build images in clouds or redistribute exported images as archive files.
But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.

@@

\label{rhub}
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \CRANpkg{rhub} \citep{csardi_rhub_2019}.
The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.

@@

There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
In general, Alpine-based Docker images are often chosen for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage.
R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
For R though, this advantage quickly deteriorates if more features are needed, e.g., if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
In fact, using Debian's 'slim' variant and removing capabilities, one could probalby achieve similarly small image sizes on a more common stack.

@@

_Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
The practice to apply a combination tools and software stacks for analysing data and the cross-cutting skillsets connected with data science puts the management of computing environments and the challenges of reproducibility high up on the agenda.
This awareness leads to a large number of bespoke Docker images but also some commonly used image stacks, ranging from base images with very common tools to 'catch all' images including _a lot_ of software.
Here we only mention a few image stacks as examples, which are not based on Rocker images or stem from the R community, but prominently include installations of R.
A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).

@@

**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
It includes machine learling libraries such as Tensorflow and Keras, and also configures the \CRANpkg{reticulate} package.
The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.

>>>>

  We identified common themes across this diversity: reproducible environments, scalability and effectivity, and portability across clouds.

@@

The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by descriptions of projects developing images with R installations.
Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers mediately or only for specific features, up to complex use cases leveraging containers.
We conclude with a reflection on the landscape of packages and applications and point out future directions of development.

```{r hexSticker, message=TRUE, warning=FALSE, cache=TRUE, include=FALSE}
if(!file.exists("rockerverse.png")) {
  library("magick")
  library("here")

  # https://commons.wikimedia.org/wiki/File:NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg
  universe <- magick::image_read("https://upload.wikimedia.org/wikipedia/commons/6/69/NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg")
  magick::image_write(universe, here("universe.png"))
  # rockerLogoMask <-
  magick::image_read("https://avatars2.githubusercontent.com/u/9100160?s=200&v=4") %>%
    magick::image_scale("2100x2100") %>%
    magick::image_channel("Matte") %>%
    image_negate() %>%
    magick::image_border("black", "150x0") %>%
    magick::image_write(here("rocker_mask.png"))

  # does not work: image_composite(universe, rockerLogoMask, operator = "CopyOpacity", compose_args = "-composite -alpha Off")
  system2("convert", c(here("universe.png"), here("rocker_mask.png"),
    "-alpha Off",
    "-compose CopyOpacity",
    "-composite",
    here("whaleverse.png")))

  library("hexSticker")
  hexSticker::sticker(here("whaleverse.png"), package = "Rockerverse",
          p_size = 20, s_x = 0.97, s_y = 0.95, s_width = 1,
          h_fill = "#000000", h_color = "#ffffff",
          url = "github.com/nuest/rockerverse-paper",
          u_color = "#ffffff", u_size = 4,
          filename = here("rockerverse.png"))

  file.remove(c(here("universe.png"), here("whaleverse.png"), here("rocker_mask.png")))
}
```

![Rockerverse hex sticker](`r here::here("rockerverse.png")`)

@@

The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
This is a rather attractive proposition and Docker's very easy to use user interface has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep[e.g.,][]{Bernstein2014}, data science \citep[e.g.,][]{boettiger_introduction_2015}, and edge computing \citep[e.g.,][]{alam_orchestration_2018}.
It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
Images are normally build from plain text documents called `Dockerfile`.
A `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.

For statistical computing and analysis centered around R, the Rocker&nbsp;Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e. smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

## Images for alternative R distributions
\label{alternatives}

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches or provide features different to what _GNU-R_, or Base R, provides.
GNU-R is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/).
While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to others.
All these distributions of R are published under GPL licenses.
Since all of the R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.

**Microsoft R Open** (MRO) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g. the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf. [http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
Packages are loaded on demand, i.e., at the first call to `library()`.
Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.

@@

**FastR** ([https://github.com/oracle/fastr](https://github.com/oracle/fastr)) is "A high-performance implementation of the R programming language, built on GraalVM" \citep{oracle_labs_oraclefastr_2020}.
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g. [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).

@@

_Bioconductor_ ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
_Bioconductor_ has a semiannual release cycle, each release is associated with a particular version of R.
Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
_Bioconductor_'s 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio` images.
_Bioconductor_ installs packages based on the R version, and therefore uses `rocker/rstudio` and `rocker/r-ver` version tagging.
_Bioconductor_ selects the desired version of R from Rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
The images are summarized on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub ([https://github.com/Bioconductor/bioconductor_docker](https://github.com/Bioconductor/bioconductor_docker)), and [distributed through DockerHub](https://hub.docker.com/u/bioconductor).
Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific image tag.
A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
Images may be valuable in teaching, where participants pull pre-built images to avoid complicated configuration of their own computing environemnts.

_Bioconductor_ has several images in addition to 'base', specific to various areas of research.
The 'core' image installs the most commonly used _Bioconductor_ packages.
_Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
A recent audit of the _Bioconductor_ `Dockerfile`s, following best practices from the Docker website, led to a reduction in the size and number of layers.
The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
A subsequent layer that might appear to remove or overwrite the path actually masks, rather than alters, the original.
It is therefore important to clean up (e.g., cache removal) within each layer, and to avoid re-installing existing dependencies.

@@

**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
All these images can be run on Docker for Desktop on Windows 10.
New R versions are only added on demand.
The explorative rocker-win project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
The latter is quite comfortable and fast actually, compared to installation from source on Linux, as R ships pre-compiled binaries for Windows.
However, compared to the various Linux images, the footprint of Windows images is quite large.

For Windows, just as originally for Linux, the cloud use cases drive the development and Docker and Microsoft collaborate closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
Naturally, a Windows license is required for the host machine, and the licensing impacts the potnetial to build images in clouds or redistribute exported images as archive files.
But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.

@@

\label{rhub}
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \pkg{rhub} \citep{csardi_rhub_2019}.
The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.

@@

There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
In general, Alpine-based Docker images are often chosen for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage.
R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
For R though, this advantage quickly deteriorates if more features are needed, e.g. if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
In fact, using Debian's 'slim' variant and removing capabilities, one could probalby achieve similarly small image sizes on a more common stack.

@@

_Data Science_ is a widely discussed topic among all academic disciplines \citep[cf. e.g.,][]{donoho_50_2017}.
Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
The practice to apply a combination tools and software stacks for analysing data and the cross-cutting skillsets connected with data science puts the management of computing environments and the challenges of reproducibility high up on the agenda.
This awareness leads to a large number of bespoke Docker images but also some commonly used image stacks, ranging from base images with very common tools to 'catch all' images including _a lot_ of software.
Here we only mention a few image stacks as examples, which are not based on Rocker images or stem from the R community, but prominently include installations of R.
A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).

@@

**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
